BivData Setup

### 2020-01-24 doc/Bivalent_Analysis.Rmd
git status

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
#library(pwr)
library(plyr)
library(lattice)
library(dplyr)
library(raster)
library(lme4)
library(nlme)
library(RLRsim)
library(reshape2)

@@

This plot illustrates how well the algorithm seperated indiviual bivalents per cell image. 
Several MSM male spreads had tight cells with a high degree of overlap, so fewer bivalents were indiviudally segmented.

```{r tech.ect, echo=FALSE, eval=TRUE}

@@

#This isn't the same as curated/passing bivalents

@@

consider making a table with predictions for bivalent patterns which are related to the mechanism

@@

Should I be thinking of making a MM or lm for predicting CO number (within a category)

but there is so little variance in the CO number

co ~ sc.length

I ran this model before,  I remember that these models are better predictors (more significant for males than females)

maybe -- female nuanced pattern -- falls along the lines of

simple predictions not met as strongly as in male patterns...

this reasoning might be complicated by the fact that the range in CO averages is lower in females -- so there is less power

@@

the above plot displays the average proportion of each type of chromosome class across mice. The results mostly align with the hand measured results.

Look into ordering the mouse strains to highlight the MSM and PWD differences, renameing the chrm type labels, spacing out the x axis labels or removing some of the other mice.

# Chromosome Size effect

In an attempt to deal with the chromosome effect size that is know to effect total rate and CO position, we try to make a dataset of the longest 25% SC from a cell.

The majority of cells from the BivData set do not have the complete completement of bivalents, (ie they are missing up to 5) -- which makes it hard to use a absolute length threshold since some cells have shorter or longer bivalents compared to the while. So an absolute measure might catch the lower 25% of some cells.

I'd like to figure out how to best seperate out the true 4th quartille bivalents (longest 25%), even with this missing data. double check that 4th quartille is top 5/4

Could I use the total SC? (I would be warry: can't ensure that they apply the same skeletonize algorithm/parameters and the XY is also included in that count and will mess everything up.)

outline logic for filtering / choosing BivData set of almost whole cells

- 16 to 20 bivalents in the cells

```{r chrm.size.effect, echo=FALSE}

@@

  facet_wrap(~strain, scales="free")+ylim(c(30,150))+ggtitle("Male Dom whole cell manual measures")
show.the.cells_male_dom

@@

  facet_wrap(~strain, scales="free")+ylim(c(30,150))+ggtitle("Dom male, almost whole cell \nMean, median, 4th Quartille ploted")
almost_cells_Dom_m

@@

I'm purposing to take all the automated biv measures above the cell's 4th quartille as a dataset for 'longestest' bivalents which we would predict to exclude the shortest chromsomes which would have stronger chromsome size effects.

@@

Biv_Data_mean_table has the Q4 values

@@

real.long.bivData <- BivData_from_15biv_cells[BivData_from_15biv_cells$long.Biv == "yes",]
#table(real.long.bivData$fileName)

#!!! omg it finally worked!~!!#OMFG

@@

Well fixing those bigs took forever. Use the DF, **real.long.bivData** for comparing the longest bivalents.  still v unblancd across sexes and strains ~ merge with the manual whole cell biv measures later...

MSM male only has data from 1 cell, and PWD 3, uh oh -- may need to merge sooner.

First proof of principl will be comparisons within category (long vs all)

@@

in code chunk above I ran the mouse averages for the longest bivalets (680 bivalents, from 54 mice. full dataset is 10202 bivalents from 86 mice.

@@

female_long_biv_plot <- ggplot(real.long.bivData[real.long.bivData$sex == "female",], aes(x=fileName, y=chromosomeLength, color=strain))+geom_point()+facet_wrap(~category, scales="free")+ylim(c(30,200))
female_long_biv_plot

male_long_biv_plot <- ggplot(real.long.bivData[real.long.bivData$sex == "male",], aes(x=fileName, y=chromosomeLength, color=strain))+geom_point()+facet_wrap(~category, scales="free")+ylim(c(30,200))
male_long_biv_plot

@@

These plots show the SC lengths for bivalents I'm calling the 'long SC data set'. They are supposed to be the longest 4-5 SC from cells where I could get good measures. These longer bivalents are useful because their patterns shouldn't be affect by chromsome size effect (which effects, CO position). Hopefully this data set will have less noise from chromosome identity, but there was still data missing (they don't come from whole cell measures).

@@

- Of all female single bivalents 5% are XXs (1 of 20).

- The XX is large, thought to be within the top 25% longest bivalents of the cell (3rd largest by Mb).

- What is the average % of XX for whole cell SC (sum(all bivalents)). Lets guess 12% of a cell's total SC area is XX.

- do the algebra...

@@

I will use two different mixed models using lme() and lmer(), because the different models have limitations for reporting p values ect.

list out pros and cons for using mouse averages of bivalent traits instead of using the single bivalent traits them selves.

@@

@@

@@

1) SC length (cite Lynn)

2.A) normalized CO positions (sedell and Kirkpatrick).

@@

3) Interference, measured in the physical units, has been shown to lack sexual dimorphism (petkov).

These 3 hypotheses will be tested with the mixed model framework, specifically focusing on the effect of sex.

Mouse averages for the each metric will be used.

## Predictions, male polymorphism

What predictions do I have for how traits will differ between the low and high Musc mouse strains?

A) **SC lengths** will be longer for high Rec

B.1) **Interfernce/IFD** will be shorter in high Rec

or

B.2) **Interference/IFD** will be longer in high Rec males, sis-co-ten metric will be maximized. (given prelim results from Beth's data)

C) It's unknown if **1CO normalized positions** between high and low musc strains, will differ. (no good predictions come to mind)

D) Sis-co-ten predictions fall into

@@

```{r MM_2_SC.length, echo=FALSE}
#Long_biv_mouse.avs_4MM

#as.data.frame( round(  summary(BivData.MixedModel.HQ)$tTable, 6 ) )

#can't make the 
SC_length.lmer.MM <- lmer(mean_SC ~ subsp * sex + (1|strain), data=mouse.avs_4MM, REML = TRUE)

anova(SC_length.lmer.MM)

thing1  <- summary(SC_length.lmer.MM)
thing2 <- drop1(SC_length.lmer.MM, test="Chisq")

@@

##long bivalents
long.bivs_SC_length.lmer.MM <- lmer(mean_SC ~ subsp * sex + (1|strain), data=Long_biv_mouse.avs_4MM, REML = TRUE)

anova(long.bivs_SC_length.lmer.MM) 
#sex is larges explantion of variance

thing1  <- summary(long.bivs_SC_length.lmer.MM)
thing2 <- drop1(long.bivs_SC_length.lmer.MM, test="Chisq")

rand_SC.Length_MM <- exactRLRT(long.bivs_SC_length.lmer.MM)
rand_SC.Length_MM

#strain no sig, but underpowered

```{r justify.mouse.level, echo=FALSE}

@@

#

#justify mouse averages
#is sampling -- bivalent dataset 
# random samples 
#equivalent sample size

@@

The exactRLRT() test indicates that the random strain effect might be significant, p= `r rand_SC.Length_MM$p.value`.

@@

```{r MM.IFD, echo=FALSE}

IFD_ABS.MM <- lmer(mean_IFD.2CO_ABS ~ subsp * sex + (1|strain), data=mouse.avs_4MM, REML = TRUE)

@@

library(lmerTest)
#anova(CO_position.MM)
anova(CO_position.MM, verbose = TRUE)

@@

#as.data.frame( round(  summary(CO_position.MM)$tTable, 6 ) )  #lme() needed for this

#thing2 <- drop1(SC_length.MM, test="Chisq")

CO_pos_rand <- exactRLRT(CO_position.MM)
CO_pos_rand

@@

The metric Sis-co-ten measures the amount of sister cohesion connected to the other pole. Alternating IFDs. Assuming chromatid interference and no sister seperation, this should reflect the amount of sister cohesion that is pulled across the opposite pole.

The logic of how the sis-co-ten metric is outlined in the figure below. The authors from this paper call it by something else. looking at this siscoten metric is a way to model different cohesin outcomes as a consequence of different numbers and placements of chiasmata/CO. This metric is calculated from SC area -- which we are using as a proxy to the amount of cohesion at metaphase.

![Logic of Sis-co-ten metric, from (Lee, J. (2019). Is age-related increase of chromosome segregation errors in mammalian oocytes caused by cohesin deterioration?. Reproductive Medicine and Biology.)](figureFiles/Lee2019_Fig2.png)

@@

Curated_BivData <- Curated_BivData[Curated_BivData$hand.foci.count < 5,]

sis.co.ten_with.plot <- ggplot(data = Curated_BivData[Curated_BivData$category == "PWD female",], aes(y=SisCoTen, x=hand.foci.count, color = as.factor(hand.foci.count) ))+
  geom_jitter()+facet_wrap(sex~strain)
sis.co.ten_with.plot

sis.co.ten_with.plot_2 <- ggplot(data = Curated_BivData[Curated_BivData$category == "PWD male",], aes(y=SisCoTen, x=hand.foci.count, color = as.factor(hand.foci.count) ))+
  geom_jitter()+facet_wrap(sex~strain)
sis.co.ten_with.plot_2

@@

Males have much clearer seperation of siscoten across chrm classes. This is emphasized when SC length is also plotted.

>>>>

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
#library(pwr)
library(plyr)
library(lattice)
library(dplyr)
library(raster)
library(lme4)
library(nlme)
library(RLRsim)
library(reshape2)
#library(lmerTest) #loading this package will mess up RLRsim's exactRTRL()
#for got the function i wanted to use

@@

Bivalent level traits and metrics have been added in the src/Setup_BivData.Rmd script. These observations are from the automated image analysis algorithm and have been currated (threw out incorrect algorithm output).

The breakdown of curated bivalents by category

`r data.frame(table(droplevels(Curated_BivData$category) ) )`

```{r tech.ect, echo=FALSE, eval=FALSE}
#this fig not needed

@@

#This isn't the same as curated/passing bivalents           
#This plot illustrates how well the algorithm seperated indiviual bivalents per cell image. 
#Several MSM male spreads had tight cells with a high degree of overlap, so fewer bivalents were indiviudally #segmented.

@@

@@

The figure below shows the average percentages of chromosome classes per category.

@@

These results align with the previous figure I made using percentages from hand measured results (dozens of observations, instead of hundreds). These results fit with the rapid male evolution of gwRR in MSM and PWD.

- Look into ordering the mouse strains to highlight the MSM and PWD differences, renameing the chrm type labels, spacing out the x axis labels or removing some of the other mice.

# Chromosome Size Effect

In an attempt to test for and metigate noise based on chromosome identity/physical size, I tried to isolate bivalents which are in the top quartille for SC length from their cells.

Since the majority of cells from the Automated BivData set do not have the complete completement of bivalents this subset is much smaller and might have 'smaller' bivalents from cells missing the longer bivalents. But assuming the algorithm's identification of bivalents in random, with a large enough sample size, hopefully this error would be minimized.

Below are examples of plots which show the SC length distributions across cells. The top figure shows whole cell hand measured data and the bottom shows the Automated bivData from cells with at least 15 bivalents measured. There are plot pairs for each category, but they are excluded for space.

Each point is a bivalent ploted by cell on the x axis. X's are the 4th quartile, big point is the mean and smaller black point is the median. I'm using these to compare the patterns of these statistics in the automated dataset which is missing some bivalent data.

```{r SC.distb_manual_Auto, echo=FALSE}

@@

  facet_wrap(~strain, scales="free")+ylim(c(30,150))+ggtitle("Male Dom whole cell manual measures")

@@

  facet_wrap(~strain, scales="free")+ylim(c(30,150))+ggtitle("Dom male, almost whole cell \nMean, median, 4th Quartille ploted")

@@

I'm purposing to take all the automated biv measures above the cell's 4th quartille as a dataset for 'longest' bivalents which we would predict to exclude the shortest chromsomes which would have stronger chromsome size effects.

@@

@@

real.long.bivData <- BivData_from_15biv_cells[BivData_from_15biv_cells$long.Biv == "yes",]
#table(real.long.bivData$fileName)
#!!! omg it finally worked!~!!#OMFG

@@

The DF **real.long.bivData** contains `r length(real.long.bivData$Obj.ID)` bivalent measures. full dataset is 
`r length(Curated_BivData$Obj.ID)`. This is the breakdown of bivalent observations by category `r data.frame(table(droplevels(real.long.bivData$category)))`.

Still very unbalanced across sexes and strains ~ merge with the manual whole cell biv measures later...
Musc males are very under-represented. MSM male only has data from 1 cell.

- Try to merge this DF with the whole.cell manual measures.

- Try estimating which 'loose' bivalent observations might be within the long class of bivalents.

@@

in code chunk above I ran the mouse averages for the longest bivalets. (680 bivalents, from 54 mice. 
10202 bivalents from 86 mice.

@@

female_long_biv_plot <- ggplot(real.long.bivData[real.long.bivData$sex == "female",], aes(x=fileName, y=chromosomeLength, color=strain))+geom_point()+facet_wrap(~category, scales="free")+ylim(c(30,200))+ggtitle("Long Bivalent Dataset, \nFemale")
female_long_biv_plot

male_long_biv_plot <- ggplot(real.long.bivData[real.long.bivData$sex == "male",], aes(x=fileName, y=chromosomeLength, color=strain))+geom_point()+facet_wrap(~category, scales="free")+ylim(c(30,200))+ggtitle("Long Bivalent Dataset, \nmale")
male_long_biv_plot

@@

These plots show the SC lengths for the 'long SC data set'. They are supposed to be the longest 4-5 SC from cells where I could get good measures. These longer bivalents are useful because their patterns shouldn't be affect by chromsome size effect (which effects, CO position). Hopefully this data set will have less noise from chromosome identity, but there was still data missing (they don't come from whole cell measures).

@@

- Of all female single bivalents observations, 5% are XXs (1 of 20).

- The XX is large, likely within the top 25% longest bivalents of the cell (3rd largest by Mb).

- The average % of XX for whole cell SC (sum(all bivalents)) can be calculated from the whole.cell dataset. Lets guess 12% of a cell's total SC area is XX.

-

@@

I will use two different mixed models using lme() and lmer(), because the different models have limitations for reporting p values for fixed and random effects.

Pros and con list for using mouse averages vs single bivalent traits.

@@

The mouse averages for all the metrics is what is used in the mixed models.

@@

                       #CO position things, requires the melt?
              norm.first.foci.pos =  mean(PER_Foci_1[which(hand.foci.count == 1)]),

              #sis.co.ten!!
              mean.siscoten = mean(SisCoTen, na.rm=TRUE)
)

@@

1) SC length will be sexually dimorphic (sex effect will be signicant)(cite Lynn)

2.A) Normalized 1CO positions will be sexually dimorphic (sedell and Kirkpatrick).

@@

  C) Centromere and telomere distances will be sexually dimorphic.

3) Interference / IFD will not be sexually dimorphic. Previous measures has been shown to lack sexual dimorphism (cite petkov).

## Predictions, Male Polymorphism (High vs Low Rec strains)

A) **SC lengths** will be longer for high Rec strains.

B.1) **Interfernce/IFD** will be shorter in high Rec strains.

C.1) It's unknown if **1CO normalized positions** between high and low musc strains, will differ. (no good predictions come to mind)

C.2) sis-co-ten metric will be maximized in higher recombining strains .. because.

C.3) telomere and centromere distances will be shorter in low rec strains

@@

```{r MM_2_SC.length, echo=FALSE, eval=FALSE}
#Long_biv_mouse.avs_4MM
#as.data.frame( round(  summary(BivData.MixedModel.HQ)$tTable, 6 ) )

#did model isn't working
SC_length.lmer.MM <- lmer(mean_SC ~ subsp * sex + (1|strain), data=mouse.avs_4MM, REML = TRUE)

#anova(SC_length.lmer.MM)
#thing1  <- summary(SC_length.lmer.MM)
#thing2 <- drop1(SC_length.lmer.MM, test="Chisq")

@@

##long bivalents
#long.bivs_SC_length.lmer.MM <- lmer(mean_SC ~ subsp * sex + (1|strain), data=Long_biv_mouse.avs_4MM, REML = TRUE)

#anova(long.bivs_SC_length.lmer.MM) 
#sex is larges explantion of variance
#thing1  <- summary(long.bivs_SC_length.lmer.MM)
#thing2 <- drop1(long.bivs_SC_length.lmer.MM, test="Chisq")

#rand_SC.Length_MM <- exactRLRT(long.bivs_SC_length.lmer.MM)
#rand_SC.Length_MM
#strain no sig, but underpowered
```

```{r justify.mouse.level, echo=FALSE, eval=FALSE}

@@

#
#justify mouse averages
#is sampling -- bivalent dataset 
# random samples 
#equivalent sample size

@@

The exactRLRT() test indicates that the random strain effect might be significant, p= xx

@@

```{r MM.IFD, echo=FALSE}
#ugg what changed, FFF kntir is broken!!

library(RLRsim)
IFD_ABS.MM <- lmer(mean_IFD.2CO_ABS ~ subsp * sex + (1|strain), data=mouse.avs_4MM, REML = TRUE)

@@

#library(lmerTest)
#anova(CO_position.MM)
anova(CO_position.MM, verbose = TRUE)

@@

#as.data.frame( round(  summary(CO_position.MM)$tTable, 6 ) )  #lme() needed for this
#thing2 <- drop1(SC_length.MM, test="Chisq")

CO_pos_rand <- exactRLRT(CO_position.MM)
CO_pos_rand
```

@@

The metric Sis-co-ten measures the amount of sister cohesion connected to the other pole.

The logic of how the sis-co-ten metric is outlined in the figure below. The goal is to use this metric to model different tension ative cohesin amounts as a consequence of different numbers and placements of chiasmata/CO. This metric is calculated using SC area as a proxy to the amount of cohesion at metaphase.

![from (Lee, J. (2019). Is age-related increase of chromosome segregation errors in mammalian oocytes caused by cohesin deterioration?. Reproductive Medicine and Biology.)](figureFiles/Lee2019_Fig2.png)

@@

jitter_siscoten_PWDf <- ggplot(data = Curated_BivData[Curated_BivData$category == "PWD female",], aes(y=SisCoTen, x=hand.foci.count, color = as.factor(hand.foci.count) ))+ ggtitle("PWD female Sis.co.ten")+
  geom_jitter()+facet_wrap(sex~strain)
jitter_siscoten_PWDf

jitter_siscoten_PWDm <- ggplot(data = Curated_BivData[Curated_BivData$category == "PWD male",], aes(y=SisCoTen, x=hand.foci.count, color = as.factor(hand.foci.count) ))+ggtitle("PWD male sis.co.ten")+
  geom_jitter()+facet_wrap(sex~strain)
jitter_siscoten_PWDm

jitter_siscoten_WSBf <- ggplot(data = Curated_BivData[Curated_BivData$category == "WSB female",], aes(y=SisCoTen, x=hand.foci.count, color = as.factor(hand.foci.count) ))+ggtitle("WSB female sis.co.ten")+
  geom_jitter()+facet_wrap(sex~strain)
jitter_siscoten_WSBf

jitter_siscoten_WSBm <- ggplot(data = Curated_BivData[Curated_BivData$category == "WSB male",], aes(y=SisCoTen, x=hand.foci.count, color = as.factor(hand.foci.count) ))+ggtitle("WSB male sis.co.ten")+
  geom_jitter()+facet_wrap(sex~strain)
jitter_siscoten_WSBm
```

Males have much clearer seperation of siscoten across chrm classes. This is emphasized when SC length is also plotted. It seems like musc males have higher amounts of this metric compared to Dom males (which might)

To formally test the differences in sis-co-ten I plan to write a subsampling / permutation loop to compare the mean(sis.co.ten) of the same numbers of bivalents of the same class.

```{r sis.co.ten.test_HetC, echo=FALSE}
#mouse.avs_4MM  - mouse averages with mean sis.co.ten

#MM
lme_MM_siscoten <- lme(mean.siscoten ~ subsp * sex, data= mouse.avs_4MM, random=list(strain=pdDiag(~sex) ) )
as.data.frame( round(  summary(lme_MM_siscoten)$tTable, 6 ) )

#sex and sex*subsp interaction effect are significant for mean sis.co.ten

lmer_MM_siscoten <- lmer(mean.siscoten ~ subsp * sex + (1|strain), data=mouse.avs_4MM, REML = TRUE)
#anova(lmer_MM_siscoten)

#thing1  <- summary(lmer_MM_siscoten)
#drop1(lmer_MM_siscoten, test="Chisq")

rand_siscoten_MM <- exactRLRT(lmer_MM_siscoten)
rand_siscoten_MM
#slightly sig
```

The fixed effects, sex and sex*subsp are significant. The random strain effect is also significant.

> Is the heterochiasmy prediction met?

Yes, model od predicting the mouse average siscoten, sex and sex-subp interaction are significant factors. The Random strain effect is also significant.

```{r siscoten.test.poly, echo=FALSE}

mouse.avs_4MM$Rec.group <- ifelse(grepl("PWD", mouse.avs_4MM$strain), 1, 
                                    ifelse(grepl("MSM", mouse.avs_4MM$strain), 1, 0))

#test high vs low Musc strains
mouse.avs_4MM

@@

> is the male polymorphism

### 2020-01-24 src/BivData_Setup.Rmd
git status

add sis_co_ten  ref Lee 2019

@@

                       mean.SC_length = mean(chromosomeLength, na.rm = TRUE),
                       var.SC_length = var(chromosomeLength, na.rm = TRUE),
                       sd.SC_length   = round(sd(chromosomeLength), 3),
                       se.SC_length   = round(sd.SC_length / sqrt(nbivs), 3)
)

>>>>

add mean.telomere, mean cent.dist and mean sis.coten to the mouse level table

update IFD functions, to add IFD2,3

@@

                       mean.SC_length = mean(chromosomeLength, na.rm = TRUE),
                       var.SC_length = var(chromosomeLength, na.rm = TRUE),
                       sd.SC_length   = round(sd(chromosomeLength), 3),
                       se.SC_length   = round(sd.SC_length / sqrt(nbivs), 3),

                       #sis.co.ten
                       mean.sis.co.ten = mean(SisCoTen, na.rm = TRUE)
)
